---
title: "RR Assignment"
output: html_document
---
```

```{r echo=TRUE}
install.packages("ggplot2")
library(ggplot2)
install.packages("magrittr")
library(magrittr)
install.packages("dplyr")
library(dplyr)
```
##Loading and pre-processing data

Used Rstudiocloud.Therefore converted csv file into an excel file and imported the dataset.
```{r echo=TRUE}
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile="./data/activity.zip",method="curl")

unzip(zipfile="./data/activity.zip",exdir="./data")
activity <- read.csv("./data/activity.csv")
activity$date <- as.Date(activity$date)
```

##Mean number of steps taken per day
```{r echo=TRUE}
stepsPerDay <- activity %>%
        group_by(date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 
head(stepsPerDay,10)
```

##Histogram
```{r echo=TRUE}
hist(stepsPerDay$sumsteps, main = "Histogram of Daily Steps", 
     col="blue", xlab="Steps", ylim = c(0,30))
```

##Mean and median number of steps taken per day
```{r echo=TRUE}
meanPreNA <- round(mean(stepsPerDay$sumsteps),digits = 2)
medianPreNA <- round(median(stepsPerDay$sumsteps),digits = 2)

print(paste("The mean is: ", meanPreNA))
print(paste("The median is: ", medianPreNA))
```

##Average daily activity pattern

1 Time series pattern
```{r echo=TRUE}
stepsPerInterval <- activity %>%
        group_by(interval) %>%
        summarize(meansteps = mean(steps, na.rm = TRUE)) 
head(stepsPerInterval,10)
plot(stepsPerInterval$meansteps ~ stepsPerInterval$interval,
     col="blue", type="l", xlab = "5 Minute Intervals", ylab = "Average Number of Steps",
     main = "Steps By Time Interval")
```

2 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE}
print(paste("Interval containing the most steps on average: ",stepsPerInterval$interval[which.max(stepsPerInterval$meansteps)]))

print(paste("Average steps for that interval: ",round(max(stepsPerInterval$meansteps),digits=2)))
```
##Imputing missing values

1 Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r echo=TRUE}
print(paste("The total number of rows with NA is: ",sum(is.na(activity$steps))))
```

2 Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
NA Strategy - To solve for the missing NA values the average for the associated interval will be used. The average was built in a prior step and is readily available: First, loop through all records of a copy of the ‘activity’ data. Then, look for records containing NA values. Finally, transform the ‘steps’ value based on matching the interval in the ‘stepsPerInterval’ data frame created in a prior step.

3 Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{R echo=TRUE}
activityNoNA <- activity  
for (i in 1:nrow(activity)){
        if(is.na(activity$steps[i])){
                activityNoNA$steps[i]<- stepsPerInterval$meansteps[activityNoNA$interval[i] == stepsPerInterval$interval]
        }
}
head(activityNoNA,10)
```

4 Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r echo=TRUE}
stepsPerDay <- activityNoNA %>%
        group_by(date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 
head(stepsPerDay,10)
hist(stepsPerDay$sumsteps, main = "Histogram of Daily Steps", 
     col="blue", xlab="Steps")
```

```{r echo=TRUE}
meanPostNA <- round(mean(stepsPerDay$sumsteps), digits = 2)
medianPostNA <- round(median(stepsPerDay$sumsteps), digits = 2)

print(paste("The mean is: ", mean(meanPostNA)))
print(paste("The median is: ", median(medianPostNA)))
```

## Are there anyDifferences in activity patterns between weekdays and weekdays

1 Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r echo=TRUE}
activityDoW <- activityNoNA
activityDoW$date <- as.Date(activityDoW$date)
activityDoW$day <- ifelse(weekdays(activityDoW$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
activityDoW$day <- as.factor(activityDoW$day)
```

2 Make a panel plot containing a time series plot (i.e. type=“l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r echo=TRUE}
activityWeekday <- filter(activityDoW, activityDoW$day == "weekday")
activityWeekend <- filter(activityDoW, activityDoW$day == "weekend")

activityWeekday <- activityWeekday %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
activityWeekday$day <- "weekday"

activityWeekend <- activityWeekend %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
activityWeekend$day <- "weekend"

wkdayWkend <- rbind(activityWeekday, activityWeekend)
wkdayWkend$day <- as.factor(wkdayWkend$day)


g <- ggplot (wkdayWkend, aes (interval, steps))
g + geom_line() + facet_grid (day~.) + 
        theme(axis.text = element_text(size = 12),axis.title = element_text(size = 14)) + 
        labs(y = "Number of Steps") + labs(x = "Interval") + 
        ggtitle("Average Number of Steps - Weekday vs. Weekend") + 
        theme(plot.title = element_text(hjust = 0.5))
```

The chart shows slight differences in the step patterns throughout the average daily intervals. For weekdays there is a larger spike in the morning intervals that could indicate commuters on their way to work. Spikes during the late morning and early afternoon seem to be higher on the weekends. One can assume that this might be related to subjects running errands, doing yardwork, exercising, etc. Those same intervals for weekdays appear to be less perhaps due to subjects sitting at their desk, driving a truck or standing in front of a machine in a factory.





